\section{Day 22: Local Extrema (Oct 23, 2025)}

\begin{definition}[Local Max]
    Have $f : A \subseteq \mathbb{R}^n \to \mathbb{R}$, with $a \in A$. $f$ has a local maximum at $a$ if there exists a $\delta > 0$, if $x \in A \cap B_ \delta(a)$ then $f(x) \leq f(a)$.
\end{definition}

\noindent The definition for local min is similar, except with $f(x) \geq f(a)$. A local extremum is a local max or min. A point $a$ can be both a local max and local min, such as when the function is constant on $B_ \delta(a)$.

\begin{simplethm}[Single Var Local EVT]
    Let $I \subseteq \mathbb{R}$ be open, with $c \in I$. If $f$ has a local extremum at $c$ then $f'(c) = 0$, or $f$ is not differentiable at $c$. 
\end{simplethm}

\begin{proof}
    Suppose $f$ has a local maximum at $c$. Then there exists a $\delta$, such that for all $x \in I \cap B_ \delta(c)$, $f(x) \leq f(c)$. Take $a, b \in B_ \delta(c)$, $a < c < b$. Then $\frac{f(a) - f(c)}{a - c}$ is some non-negative value. and $\frac{f(b) - f(c)}{b - c}$ is some non-positive value. Suppose $f$ is differentiable at $c$. Then taking the limit to $c$ from the left hand side, get
    \[
    f'(c) = \lim_{ x \to c^- } \frac{f(x) - f(c)}{x - c} \geq 0
    \]
    and $f'(c) \leq 0$ for the right hand limit as well. Hence $f'(c) = 0$. 
\end{proof}

\begin{simplethm}[Local EVT] \label{thm:localevt}
    Let $A \subseteq \mathbb{R}^n$, $f : A \to \mathbb{R}$ be a real-valued function. If $a$ is an interior point of $A$, and $f$ has a local maximum at $a$, then $\nabla f(a) = 0$ or $\nabla f(a)$ does not exist.
\end{simplethm}

\begin{proof}
    Suppose the preconditions are true. It suffices to show that if $\nabla f(a)$ exists, then $\nabla f(a) = 0$. We have
    \[
    \nabla f(a) = \begin{pmatrix}
        \partial_1 f(a) \\ \cdots \\ \partial_n f(a)
    \end{pmatrix}
    \]
    Let $\delta > 0$, such that for $x \in A \cap B_ \delta(a)$, $f(x) \leq f(a)$. For $0 < h < \delta$, have
    \[
    \frac{f(a + he_i) - f(a)}{h} \leq  \frac{f(a) - f(a)}{h} = 0
    \]
    similarly for $-\delta < h < 0$, have
    \[
    \frac{f(a + he_i) - f(a)}{h} \geq \frac{f(a) - f(a)}{h} = 0
    \]
    which proves the claim.
\end{proof}

